---
title: "Using the nwos package"
author: "Brett J. Butler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---


This vignette shows the basic steps necessary to generate estimates using the nwos package using the nwos.data and nwos.data.sample datasets that are supplied with the package.

The basic steps are:
* Load the nwos package
  + The nwos package can be downloaded from GitHub (devtools::install_git("git://github.com/bbutler01/nwos.git"))
* Load data
* Calculate response rates
* Calcuate weights
* Generate estimates
* Generate estimate sampling errors

The nwos package generates estimates for totals, proportions, means, and associated sampling errors.

## General setup: Install/load packages
```{r setup}
# devtools::install_git("git://github.com/bbutler01/nwos.git")
library(nwos)
library(ggplot2)
```

## Load data
```{r load data}
load("../data/nwos_data.RData")
load("../data/nwos_data_sample.RData")
```

Datasets need to have, at a minimum, the following variables:
* OWN_ID
* POINT_COUNT
* ACRES_FOREST

Optional variables include:
* SAMPLE
* RESPONSE (1=response, 0=nonresponse, NAs are permissable)

## Calculate response rates
Response rates need to be calculated for each state and stratum (i.e., forest ownership group) for which estimates are desired. This vingette assumes the data are from one state and there is only one stratum. Additional states and stratum are calculated by looping through the approriate levels.

For this vignette, all ownerships sampled are assumed to have responded ($RR_S=1.00$)

```{r response rates}
rr <- function(s, df=nwos.data.sample)
{
  rr <- sum(df$POINT_COUNT[df$RESPONSE%in%c(1) & df$SAMPLE%in%s],na.rm=T) / 
    sum(df$POINT_COUNT[df$RESPONSE%in%c(1,0) & df$SAMPLE%in%s],na.rm=T)
  return(rr)
}
sample.list <- unique(nwos.data.sample$SAMPLE)
response.rate <- data.frame(SAMPLE=sample.list,
                            RESPONSE_RATE=sapply(unique(nwos.data.sample$SAMPLE),rr))
```

## Generate weights
```{r}
nwos.data.sample$WEIGHT <- NA
for(i in sample.list) { # By sample
  nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i] <- nwosWeights(point.count=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])),
                                                                     area=nwos.data.sample$ACRES_FOREST[nwos.data.sample$SAMPLE==i],
                                                                     stratum=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])),
                                                                     area.state=1000,
                                                                     response.rate=1)
}
```

## Generate estimates - ownership totals
```{r}
sample.total.own <- numeric(0)
for(i in sample.list) { # By sample
  sample.total.own <- c(sample.total.own,
                        nwosTotal(weight=nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i], 
                                  point.count=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                  domain=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                  area=nwos.data.sample$ACRES_FOREST[nwos.data.sample$SAMPLE==i]))
}

```

```{r}
total.own <- NROW(nwos.data)
total.own.t <- t.test(sample.total.own,mu=total.own)
```

The mean estimated number of ownerships is `r round(mean(sample.total.own),2)` with a sampling error of XX. The actual number of ownerships in the population is `r total.own`. The difference between these values is `r round(abs(mean(sample.total.own)-total.own),2)` with a t-statstic of `r round(total.own.t$statistic,2)` (p-value of `r  round(total.own.t$p.value,2)` with `r  total.own.t$parameter` degrees of freedom).

```{r}
ggplot() + geom_boxplot(aes(x="1",y=sample.total.own)) + geom_hline(aes(yintercept=total.own, color="red"))
```

## Generate estimates - acreage totals
```{r}
sample.total.area <- numeric(0)
for(i in sample.list) { # By sample
  sample.total.area <- c(sample.total.area,
                         nwosTotal(weight=nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i], 
                                   point.count=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                   domain=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                   area=nwos.data.sample$ACRES_FOREST[nwos.data.sample$SAMPLE==i],
                                   units="area"))
}
```

```{r}
total.area <- sum(nwos.data$ACRES_FOREST)
# total.area.t <- t.test(sample.total.area,mu=total.area)
```

The mean estimated area is `r round(mean(sample.total.area),2)` with a sampling error of XX. The actual area in the population is `r total.area`. The difference between these values is `r round(abs(mean(sample.total.area)-total.area),2)`.

## Estimate proportion (Y_1)
```{r}
# x <- nwos.data.sample[nwos.data.sample$SAMPLE==1,]
sample.prop.own.y1 <- numeric(0)
sample.prop.var.own.y1 <- numeric(0)
sample.prop.area.y1 <- numeric(0)
sample.prop.var.area.y1 <- numeric(0)
for(i in sample.list) { # By sample
  sample.prop.own.y1 <- c(sample.prop.own.y1,
                          nwosProportion(weight=nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i], 
                                         point.count=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                         domain=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                         y=nwos.data.sample$Y_1[nwos.data.sample$SAMPLE==i],
                                         area=nwos.data.sample$ACRES_FOREST[nwos.data.sample$SAMPLE==i]))
  sample.prop.area.y1 <- c(sample.prop.area.y1,
                           nwosProportion(weight=nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i], 
                                          point.count=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                          domain=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                          y=nwos.data.sample$Y_1[nwos.data.sample$SAMPLE==i],
                                          area=nwos.data.sample$ACRES_FOREST[nwos.data.sample$SAMPLE==i],
                                          units="area"))
}
mean(sample.prop.own.y1)
mean(nwos.data$Y_1)
mean(sample.prop.area.y1)
sum(nwos.data$Y_1*nwos.data$ACRES_FOREST) / sum(nwos.data$ACRES_FOREST)
```


## Estimate mean (size of forest holdings)

```{r}
# x <- nwos.data.sample[nwos.data.sample$SAMPLE==1,]
# nwosMean()
sample.mean.own.ac <- numeric(0)
sample.mean.area.ac <- numeric(0)
for(i in sample.list) { # By sample
  sample.mean.own.ac <- c(sample.mean.own.ac,
                          nwosMean(weight=nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i], 
                                   point.count=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                   domain=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                   y=nwos.data.sample$ACRES_FOREST[nwos.data.sample$SAMPLE==i],
                                   area=nwos.data.sample$ACRES_FOREST[nwos.data.sample$SAMPLE==i]))
  sample.mean.area.ac <- c(sample.mean.area.ac,
                           nwosMean(weight=nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i], 
                                    point.count=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                    domain=rep(1,NROW(nwos.data.sample$WEIGHT[nwos.data.sample$SAMPLE==i])), 
                                    y=nwos.data.sample$ACRES_FOREST[nwos.data.sample$SAMPLE==i],
                                    area=nwos.data.sample$ACRES_FOREST[nwos.data.sample$SAMPLE==i],
                                    units="area"))
}
mean(sample.mean.own.ac)
mean(nwos.data$ACRES_FOREST)
mean(sample.mean.area.ac)
sum(nwos.data$ACRES_FOREST*nwos.data$ACRES_FOREST) / sum(nwos.data$ACRES_FOREST)

```

