---
title: "National Woodland Owner Survey (NWOS) Raw Database Catalog"
author: "Jesse Caputo"
date: '`r format(Sys.time(),"%d %B %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # set options

#packages
library(knitr)
library(RODBC)
library(nwos)

#download data
TABLES <- sqlQuery64("SELECT * FROM FS_NWOS.TABLES")
FIELDS <- sqlQuery64("SELECT * FROM FS_NWOS.FIELDS")
CODES <- sqlQuery64("SELECT * FROM FS_NWOS.CODES")

TABLES <- TABLES[order(TABLES$TABLE_NAME),] #reorder
FIELDS <- FIELDS[order(FIELDS$TABLE_NAME,FIELDS$COLUMN),]
CODES <- CODES[order(CODES$TABLE_NAME,CODES$FIELD_NAME,as.numeric(CODES$CODE)),]
```

## Introduction

This document contains a prelimninary catalog of tables, fields, and codes in the current version of the NWOS Database.

## Tables

The DB contains `r nrow(TABLES)` tables, including administrative tables and survey tables, those tables containing information about the sampling frame, survey organization, and raw data.

```{r tables, echo=FALSE}
kable(TABLES[,c("TABLE_NAME","DESCRIPTION","TYPE")],row.names=F)
```

## Fields

The DB contains `r nrow(FIELDS)` individual fields among the `r nrow(TABLES)` tables.

```{r fields, echo=FALSE}
kable(FIELDS[,c("TABLE_NAME","FIELD_NAME","DESCRIPTION","DATA_TYPE")],row.names=F)
```

## Codes

`r length(unique(CODES$FIELD_CN))` fields in the DB have coded values.

```{r codes, echo=FALSE}
kable(CODES[,c("TABLE_NAME","FIELD_NAME","CODE","MEANING")],row.names=F)
```